Identifying Shifts in the Share Distribution of Transboundary Fish Stocks to Lead Time for Policy Response Under Climate Change

### 2020-11-13 Scripts/Manuscript.Rmd
Introduction and methods comments addressed

---
title: Identifying Shifts in the Share Distribution of Transboundary Fish Stocks to Lead Time for Policy Response Under Climate Change
subtitle: Challenges to transboundary fisheries under climate change
author: 'Authors: Juliano Palacios-Abrantes*^1^, Thomas Frölicher^2^, Gabriel Reygondeau^13^, U. Rashid Sumaila^1^, Colette Wabnitz^14^, and William W.L. Cheung^1^'
date: 'Affiliations: ^1^Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada. ^2^Climate and Environmental Physics, Physics Institute, University of Bern.'
output:
  word_document: 
    reference_docx: ../References/template.docx
  pdf_document: default
bibliography: ../References/ToE_Ref.bib
csl: ../References/science.csl
---

@@

We projected the distribution of marine species using a dynamic bioclimatic envelope model (hereafter called DBEM) [@Cheung:2010dt; @Cheung:2016jd]. The DBEM  represents species’ physiology, habitat suitability, depth and latitudinal ranges, and spatial population dynamics as well as preferences to sea temperature, salinity, oxygen content, sea ice extent (for polar species) and bathymetry. For pelagic species, the model uses environmental variables at the surface (e.g., sea surface temperature) whereas demersal species distribution are driven by ocean bottom variables (e.g., bottom temperature). This way, the model estimates species abundance and maximum catch potential (a proxy of maximum sustainable yield) over a $0.5^{\circ}$ x $0.5^{\circ}$ grid of the world (see Cheung *et al.* [-@Cheung:2010dt; -@Cheung:2016jd] for model details).

@@

$lat_{ts} = \mu (lat_{per})$

$lon_{ts} = \mu (lon_{per})$

@@

$$D_{ens} = acos(sin(lat_{eez})*sin(lat_{ts}) + cos(lat_{eez})*cos(lat_{ts})*cos(lon_{eez}-lon_{ts}))$$
where $lat_{eez}$ and $lat_{ts}$ are the latitudes of the EEZ and transboundary stock centroids, respectively and $lon_{eez}$ and $lon_{ts}$ are the respective longitudes.

@@

$$TI =(\frac{D_{A,t}}{sd(D_{A,th})} - \frac{D_{B,t}}{sd(D_{B,th)}})^2$$

where $D_A$ and $D_B$ represent the distance between the distribution centroids of a species and the centroid of the pair of neighbouring EEZs (A and B) sharing the species for each time step from 2006 to 2100 (*t*); and *sd* is the standard deviation of the historical (*th*, 1951 - 2005) centroid distribution for $D_A$ and $D_B$. Thus, a higher TI means that a shared species becomes more or less abundant relative to its neighbouring EEZs that is beyond what the fisheries experienced historically. From the perspective of a state’s fisheries, higher TI indicates an increase in the challenges to the effective management and sustainability of the shared species’ fisheries. In contrast, a smaller TI suggests that the sharing of the species between neighbouring EEZs is becoming relatively more stable; a condition that favours effective management and sustainability of the shared species’ fisheries. The index was smoothed to a 10 year average to xxx.

@@

$EmergenceTIndex = \frac{\mu_{TI[t]}}{SD_{TI_{th}}}$
$ToE = t when Emergence index[t] > 1$

where $\mu_{TI}$ is the yearly mean TI ($\mu$) across all ten ensemble members in year *yr* and $SD_{TI}$ is the standard deviation of the smoothed (10-year moving average) TI across the ensemble members for the historical reference period (1951 - 2005). We set a threshold of TI above which we consider a shared stock to have emerged from the historical internal variability of the TI. The threshold was set as one standard deviation of the variability ($SD_{TI}$) (i.e., eq. **x**, representing a probability of 68% that the transboundary index has emerged from historical variability). We tested the sensitivity of the calculated emergence index to  a higher emergence threshold of two times the standard deviation (i.e., representing a probability of 95% that the index has emerged).

@@

$$ index' = (P_c + P_v)/ToE $$

@@

First we estimated the stock share ratio (SSR) for each species and year (1951-2100) by aggregating the number of $0.5^{\circ}$ x $0.5^{\circ}$ grid cells in which the species is present across neighboring EEZs, and then calculated the proportion of the stock held within each EEZ [@PalaciosAbrantes:2020tv]. Second, we average the calculated proportion of stock within an EEZ held into three time periods to reduce the effects of inter-annual climate variability. The first time period ($t_h$) spans 1951 to 2005 and matches the historical period in the GFDLESM-2M simulations, over which the model was forced with observation-based greenhouse gas, aerosol and natural external forcing [@Dunne:2012bo; @Dunne:2013db]. It is assumed as the historic baseline of shared distributions for each species. We then selected two future periods; early $21^{st}$ century as the average of 2021 to 2040 (*te*), and mid $21^{st}$ century as the average of 2041 to 2060 (*tm*). We chose these time periods to match the challenges of achieving fisheries-related UN-SDGs [@UnitedNations2018] such as SDG 14.4 (end overfishing) and SDG 1.2 (poverty reduction), to be achieved by 2020 and 2030, respectively [@Singh:2017ds]. The analysis was replicated for projected species distributions from each of the ten ensemble members and results were averaged across ensemble members. Third, we estimated the percentage change in the SSR ($\Delta{SSR}$) of each future time period ($SSR_f$) relative to the historic time period ($SSR_h$) following Palacios-Abrantes et al. (*in press*),

$$ \Delta{SSR_{e,m}} = \frac{(SSR_{f}-SSR_{th})}{SSR_{th}}* 100 $$

$TP = \Delta{SSR}_{e,m} < μ±σ$

@@

We tested the results for normality (e.g. skewness, kurtosis) and performed a couple of non-parametric Kruskal–Wallis test by ranks [@Hollander:2013wz] to investigate geopolitical and ecological differences in the ToE of transboundary species. Specifically, we tested if the habitat association of transboundary species and the geographic location of EEZs would have any effect on the time of emergence of transboundary stocks. For both cases, our null hypothesis was that there were no significant differences in the time of emergence across habitat association nor EEZs.
<!-- Moreover, we implemented a multiple linear regression analysis to a series to explain the ToE of each species in each EEZ. Specifically, we looked at habitat preferences, catch trends, geopolitical and biome location and ToE of both bottom and sea surface temperature (See *Databases and species selection*). We hypothesize that all of these factors would have significant influence in the ToE of transboundary species. -->
All analyses were run using the statistical software *R version 3.5.2 (2018-12-20; Eggshell Igloo)* with the packages *data.table* [@Packagedatatable:2019uh], *janitor* [@PackagejanitorSim:2018ur], *rfishbase* [@PackagerfishbaseR:2019th], *sf* [@PackagesfSimpleF:2018vp], *sp* [@PackagespClasses:2019tn], *tidiverse* [@PackagetidyverseE:2017vq], *tidytext* [@PackagetidytextTe:2019vk], *gmt* [@gmt] and *zoo* [@PackagezooSInfr:2019tm], *readxl* [@REF], *ggrepel* [@REF], *zeallot* [@REF], *viridis* [@REF], *cowplot* [@REF], *moments* [@REF] and *pgirmess* [@REF]. All code is available at https://github.com/jepa/TransEmergence

@@

By 2030, changes in SSR will be seen across the globe, but with no clear pattern in the direction of change (Figure \ref{fig:Figure5}). Poleward shifts in SSR will be seen in regions of the world with relatively large EEZs and few boundaries like the North East Pacific. However, despite a general recognition that climate change is shifting species distributions poleward [@Cheung:2010dt; @Poloczanska:2016kk; @IPCC:2019tn], the Eastern Tropical Pacific will see an equatorial shift with EEZs from Mexico to Ecuador losing SSR to their southern neighbors. This shift is consistent with regional modeling studies suggesting that by mid $21^{st}$ century marine species will displace between 50 and 100 km southward of Mexico’s EEZ towards Peru under a high emission climate change scenario (Clark *et al.,* n.d.). Moreover, regions with presence of multiple frontiers like the Caribbean and the North East Atlantic will see a complex network of changes in SSR with no marked direction (Figure \ref{fig:Figure5}B). These regions are known for being highly connected by both fish adult migration [@Levin:2018in] and larvae dispersal [@Ramesh:2019va] leading to joint management plans [@Baudron:2020dj] and the call for new implemented shared policies [@Levin:2018in]. However, game theory suggests that the greater number of parties negotiating, the harder it is for parties to reach an agreement [@Sumaila:2013hv], thus, making it particularly challenging for countries in these regions to coordinate the management of shifting shared species.

>>>>

---
title: Identifying Shifts in the Share Distribution of Transboundary Fish Stocks to Lead Time for Policy Response Under Climate Change
subtitle: Challenges to transboundary fisheries under climate change
author: 'Authors: Juliano Palacios-Abrantes*^1^, Thomas Frölicher^2^, Gabriel Reygondeau^13^, U. Rashid Sumaila^1^, Colette Wabnitz^14^, and William W.L. Cheung^1^'
date: 'Affiliations: ^1^Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada. ^2^Climate and Environmental Physics, Physics Institute, University of Bern.'
output:
  word_document: 
    reference_docx: ../References/template.docx
  pdf_document: default
bibliography: ../References/ToE_Ref.bib
csl: ../References/science.csl
header-includes:
- \usepackage{amsmath}
---

@@

We projected the distribution of marine species using a dynamic bioclimatic envelope model (hereafter called DBEM) [@Cheung:2010dt; @Cheung:2016jd]. The DBEM  represents species’ physiology, habitat suitability, depth and latitudinal ranges, and spatial population dynamics as well as preferences to sea temperature, salinity, oxygen content, sea ice extent (for polar species) and bathymetry. For pelagic species, the model uses environmental variables at the surface (e.g., sea surface temperature) whereas demersal species distribution are driven by ocean bottom variables (e.g., bottom temperature). This way, the model estimates species abundance and maximum catch potential (a proxy of maximum sustainable yield) over a $0.5^{\circ}$ x $0.5^{\circ}$ grid of the world (see Cheung *et al.* [-@Cheung:2010dt; -@Cheung:2016jd] for model details). Recent work has demonstrated that, at the global scale, DBEM is able to reproduce catch changes similar in magnitude to those detected in observed catch time series in the
world’s EEZs (Cheung et al. in revision).

@@

<!-- $lat_{ts} = \mu (lat_{per})$ -->

<!-- $lon_{ts} = \mu (lon_{per})$ -->

\begin{equation}
\begin{split}
\label{eq:one}
lat_{ts} = \mu (lat_{per}) \\
lon_{ts} = \mu (lon_{per})
\end{split}
\end{equation}

@@

<!-- $$D_{ens} = acos(sin(lat_{eez})*sin(lat_{ts}) + cos(lat_{eez})*cos(lat_{ts})*cos(lon_{eez}-lon_{ts}))$$ -->

\begin{equation}
\label{eq:two}

D_{ens} = acos(sin(lat_{eez})*sin(lat_{ts}) + cos(lat_{eez})*cos(lat_{ts})*cos(lon_{eez}-lon_{ts}))

\end{equation}

where $lat_{eez}$ and $lat_{ts}$ are the latitudes of the EEZ and transboundary stock centroids, respectively and $lon_{eez}$ and $lon_{ts}$ are the respective longitudes.

@@

<!-- $$TI =(\frac{D_{A,t}}{sd(D_{A,th})} - \frac{D_{B,t}}{sd(D_{B,th)}})^2$$ -->

\begin{equation}
\label{eq:three}
TI =(\frac{D_{A,t}}{sd(D_{A,th})} - \frac{D_{B,t}}{sd(D_{B,th)}})^2
\end{equation}

where $D_A$ and $D_B$ represent the distance between the distribution centroids of a species and the centroid of the pair of neighbouring EEZs (A and B) sharing the species for each time step from 2006 to 2100 (*t*); and *sd* is the standard deviation of the historical (*th*, 1951 - 2005) centroid distribution for $D_A$ and $D_B$. Thus, a higher TI means that a shared species becomes more or less abundant relative to its neighbouring EEZs that is beyond what the fisheries experienced historically. From the perspective of a state’s fisheries, higher TI indicates an increase in the challenges to the effective management and sustainability of the shared species’ fisheries. In contrast, a smaller TI suggests that the sharing of the species between neighbouring EEZs is becoming relatively more stable; a condition that favours effective management and sustainability of the shared species’ fisheries. The index was smoothed to a 10 year average to reduce interanual variabillity.

@@

<!-- $EmergenceTIndex = \frac{\mu_{TI[t]}}{SD_{TI_{th}}}$ -->
<!-- $ToE = t when Emergence index[t] > 1$ -->

\begin{equation}
\begin{split}
\label{eq:four}
EmergenceTIndex = \frac{\mu_{TI[t]}}{SD_{TI_{th}}} \\
ToE = t when Emergence index[t] > 1
\end{split}
\end{equation}

where $\mu_{TI}$ is the yearly mean TI ($\mu$) across all ten ensemble members in year *yr* and $SD_{TI}$ is the standard deviation of the smoothed (10-year moving average) TI across the ensemble members for the historical reference period (1951 - 2005). We set a threshold of TI above which we consider a shared stock to have emerged from the historical internal variability of the TI. The threshold was set as one standard deviation of the variability ($SD_{TI}$) (i.e., Equation \ref{eq:four}, representing a probability of 68% that the transboundary index has emerged from historical variability). We tested the sensitivity of the calculated emergence index to  a higher emergence threshold of two times the standard deviation (i.e., representing a probability of 95% that the index has emerged).

@@

<!-- $$ index' = (P_c + P_v)/ToE $$ -->

\begin{equation}
\label{eq:five}
index' = (P_c + P_v)/ToE
\end{equation}

@@

First we estimated the stock share ratio (SSR) for each species and year (1951-2100) by aggregating the number of $0.5^{\circ}$ x $0.5^{\circ}$ grid cells in which the species is present across neighboring EEZs, and then calculated the proportion of the stock held within each EEZ [@PalaciosAbrantes:2020tv]. Second, we average the calculated proportion of stock within an EEZ held into three time periods to reduce the effects of inter-annual climate variability. The first time period ($t_h$) spans 1951 to 2005 and matches the historical period in the GFDLESM-2M simulations, over which the model was forced with observation-based greenhouse gas, aerosol and natural external forcing [@Dunne:2012bo; @Dunne:2013db]. It is assumed as the historic baseline of shared distributions for each species. We then selected two future periods; early $21^{st}$ century as the average of 2021 to 2040 (*te*), and mid $21^{st}$ century as the average of 2041 to 2060 (*tm*). We chose these time periods to match the challenges of achieving fisheries-related UN-SDGs [@UnitedNations2018] such as SDG 14.4 (end overfishing), SDG 2.4 (ensure sustainable food production systems) or SDG 1.2 (poverty reduction), to be achieved by 2030 [@Singh:2017ds]. The analysis was replicated for projected species distributions from each of the ten ensemble members and results were averaged across ensemble members. Third, we estimated the percentage change in the SSR ($\Delta{SSR}$) of each future time period ($SSR_f$) relative to the historic time period ($SSR_h$) following Palacios-Abrantes et al. (*in press*),

<!-- $$ \Delta{SSR_{e,m}} = \frac{(SSR_{f}-SSR_{th})}{SSR_{th}}* 100 $$ -->
<!-- $TP = \Delta{SSR}_{e,m} < \mu \pm \sigma$ -->

\begin{equation}
\begin{split}
\label{eq:six}
\Delta{SSR_{e,m}} = \frac{(SSR_{f}-SSR_{th})}{SSR_{th}}* 100 \\
TP = \Delta{SSR}_{e,m} < \mu \pm \sigma
\end{split}
\end{equation}

@@

We tested the results for normality (e.g. skewness, kurtosis) and performed a couple of non-parametric Kruskal–Wallis test by ranks [@Hollander:2013wz] to investigate geopolitical and ecological differences in the ToE of transboundary species. Specifically, we tested if the habitat association of transboundary species and the geographic location of EEZs would have any effect on the time of emergence of transboundary stocks. For both cases, our null hypothesis was that there were no significant differences in the time of emergence across habitat association nor EEZs.
<!-- Moreover, we implemented a multiple linear regression analysis to a series to explain the ToE of each species in each EEZ. Specifically, we looked at habitat preferences, catch trends, geopolitical and biome location and ToE of both bottom and sea surface temperature (See *Databases and species selection*). We hypothesize that all of these factors would have significant influence in the ToE of transboundary species. -->
All analyses were run using the statistical software *R version 3.5.2 (2018-12-20; Eggshell Igloo)* with the packages *data.table* [@Packagedatatable:2019uh], *janitor* [@PackagejanitorSim:2018ur], *rfishbase* [@PackagerfishbaseR:2019th], *sf* [@PackagesfSimpleF:2018vp], *sp* [@PackagespClasses:2019tn], *tidiverse* [@PackagetidyverseE:2017vq], *tidytext* [@PackagetidytextTe:2019vk], *gmt* [@PackagegmtInterfa:2017vh] and *zoo* [@PackagezooSInfr:2019tm], *ggrepel* [@PackageggrepelAut:2020to], *zeallot* [@PackagezeallotMul:2018wy], *viridis* [@PackageviridisDef:2018us], *cowplot* [@CowplotStreamlined:2019wt], *moments* [@PackagemomensMome:2015uh] and *pgirmess* [@PackagepgirmessSp:2018um]. All code is available at https://github.com/jepa/TransEmergence

@@

By 2030, changes in SSR will be seen across the globe, but with no clear pattern in the direction of change (Figure \ref{fig:Figure5}). Poleward shifts in SSR will be seen in regions of the world with relatively large EEZs and few boundaries like the North East Pacific. However, despite a general recognition that climate change is shifting species distributions poleward [@Cheung:2010dt; @Poloczanska:2016kk; @IPCC:2019tn], the Eastern Tropical Pacific will see an equatorial shift with EEZs from Mexico to Ecuador losing SSR to their southern neighbors. This shift is consistent with regional modeling studies suggesting that by mid $21^{st}$ century marine species will displace between 50 and 100 km southward of Mexico’s EEZ towards Peru under a high emission climate change scenario [@Clarke:2020ep]. Moreover, regions with presence of multiple frontiers like the Caribbean and the North East Atlantic will see a complex network of changes in SSR with no marked direction (Figure \ref{fig:Figure5}B). These regions are known for being highly connected by both fish adult migration [@Levin:2018in] and larvae dispersal [@Ramesh:2019va] leading to joint management plans [@Baudron:2020dj] and the call for new implemented shared policies [@Levin:2018in]. However, game theory suggests that the greater number of parties negotiating, the harder it is for parties to reach an agreement [@Sumaila:2013hv], thus, making it particularly challenging for countries in these regions to coordinate the management of shifting shared species.
